Abstraction and Invariance for Algebraically Indexed Types

### 2012-07-08 paper/section-instantiations.tex
Geometric isomorphisms

\section{Geometric Consequences of Abstraction}
\label{sec:instantiations}

\fixme{explain the general setup, with a set of operations for each
  instantiation.}

\begin{example}[Two-dimensional Geometry] We define the ``standard''
  model of the indexing theory for the two-dimensional geometry
  example as follows. Each of the sorts are interpreted just as their
  semantic counterparts:
  \begin{displaymath}
    \begin{array}{@{}llll}
      \semSort{\SynTransl{2}} = \Transl{2}
      &
      \semSort{\SynGL{2}} = \GL{2}
      &
      \semSort{\SynOrth{2}} = \Orth{2}
      &
      \semSort{\SynGL{1}} = \GL{1}
    \end{array}
  \end{displaymath}
  Each of the index operations (e.g.,~the group structure and
  determinant) are interpreted by the intended semantic
  operations. These clearly satisfy the axioms we stated in
  \exref{ex:two-dim-geo-axioms}. The assignment of relations for the
  primitive types $\tyPrimNm{vec}$ and $\tyPrimNm{real}$ are given by
  the following definitions:
  \begin{displaymath}
    \begin{array}{ll}
      R^{\bullet}_{\tyPrimNm{vec}} = \{( \vec{0}, \vec{0} ) \}&
      R_{\tyPrimNm{vec}}(B,\vec{t}) = \{ (\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \} \\
      R^{\bullet}_{\tyPrimNm{real}} = \{(0,0)\} &
      R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}
    \end{array}
  \end{displaymath}
  The definitions of $R_{\tyPrimNm{vec}}$ and $R_{\tyPrimNm{real}}$
  are as may be expected given the informal definitions we gave in
  \autoref{sec:motivating-examples}. We have set
  $R^\bullet_{\tyPrimNm{vec}}$ and $R^\bullet_{\tyPrimNm{real}}$ to
  relate $\vec{0}$ and $0$ to themselves to account for the
  polymorphic zero constants we have assumed.
\end{example}

\begin{lemma}
  The thing about $\Gamma_\Geom$.
\end{lemma}

@@

\subsection{Isomorphisms for Types Indexed by Abelian Groups}
\label{sec:types-indexed-abelian-groups}

\begin{enumerate}
\item Two cases: when we have a total ``division'' operation, and when
  we don't.
\item Type isomorphisms, based on Smith Normal Form
% \item Non-definability results, based on the non-compositional
%   relational environments
% \item Alteration of the non-definability results in the light of
%   addition of a square root operation
% \item Mention the variation we can obtain by changing some types,
%   e.g. the type of $\mathrm{abs}$. Also, the complex numbers example.
\end{enumerate}

>>>>

\section{Geometric Consequences of Abstraction}
\label{sec:instantiations}
We now instantiate the general framework of the previous section with the indexing theory
of \autoref{sec:motivating-examples}, and present
more general and formally-justified theorems for free, type isomorphisms, and non-definability
results.

%\begin{example}[Two-dimensional Geometry] 
We define the ``standard''
  model of the indexing theory for the two-dimensional geometry
  example as follows. Each of the sorts are interpreted just as their
  semantic counterparts:
  \begin{displaymath}
    \begin{array}{@{}llll}
      \semSort{\SynTransl{2}} = \Transl{2}
      &
      \semSort{\SynGL{2}} = \GL{2}
      &
      \semSort{\SynOrth{2}} = \Orth{2}
      &
      \semSort{\SynGL{1}} = \GL{1}
    \end{array}
  \end{displaymath}
  Each of the index operations (e.g.,~the group structure and
  determinant) are interpreted by the intended semantic
  operations. These clearly satisfy the axioms we stated in
  \exref{ex:two-dim-geo-axioms}. The assignment of relations for the
  primitive types $\tyPrimNm{vec}$ and $\tyPrimNm{real}$ are given by
  the following definitions:
  \begin{displaymath}
    \begin{array}{ll}
      R^{\bullet}_{\tyPrimNm{vec}} = \{( \vec{0}, \vec{0} ) \}&
      R_{\tyPrimNm{vec}}(B,\vec{t}) = \{ (\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \} \\
      R^{\bullet}_{\tyPrimNm{real}} = \{(0,0)\} &
      R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}
    \end{array}
  \end{displaymath}
  The definitions of $R_{\tyPrimNm{vec}}$ and $R_{\tyPrimNm{real}}$
  are as may be expected given the informal definitions we gave in
  \autoref{sec:motivating-examples}. We have set
  $R^\bullet_{\tyPrimNm{vec}}$ and $R^\bullet_{\tyPrimNm{real}}$ to
  relate $\vec{0}$ and $0$ to themselves to account for the
  polymorphic zero constants we have assumed.
%\end{example}

%\begin{lemma}
%  The thing about $\Gamma_\Geom$.
%\end{lemma}

@@

\subsection{Type Isomorphisms}
\label{sec:types-indexed-abelian-groups}
We say that two types $A$ and $B$ are isomorphic, and write $A \cong B$, if there exist maps
between them that are mutually inverse with respect to contextual equivalence. That is, 
there are terms $\vdash I : A \to B$ and $\vdash J : B\to A$ such
that
\[
x:A\vdash J(I(x))\stackrel{ctx}\approx x : A
\text{ and }y:B\vdash I(J(y))\stackrel{ctx}\approx y : B.
\]
It is straightforward to prove that $\cong$ is a congruence relation on types. We
can also derive isomorphisms that are independent of the indexing theory, such as
$A\times B \cong B \times A$, and $\forall i\mathord:s.A\to B\cong A\to\forall i\mathord:s.B$ for $i$ not free in $A$.

Types indexed by abelian groups induce a particularly rich theory of
type isomorphisms; previous work on units-of-measure~\cite{kennedy97relational} relates
these to Buckingham's theorem from dimensional analysis. Here we consider the additive 
abelian groups of translations and the multiplicative abelian group of scalings.

\paragraph{Translations} 
Consider first the group $\Transl{2}$ of translations as presented in
\autoref{sec:type-system-geom-intro} and \autoref{sec:affine-vector-ops}. 
Given any quantifier-free type $A$ we can prove the following isomorphism
\begin{eqnarray*}
\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{t}\to A
& \cong &
A[0/t]
\end{eqnarray*}
where $A[0/t]$ denotes substitution of the identity translation $0$
for index variable $t$ in type $A$.  Intuitively, the isomorphism
expresses the equivalence between \emph{coordinate-free} geometry, as
expressed by a polymorphic type whose first argument can be thought of
as representing the \emph{origin} with respect to which $A$ is
specified, and a \emph{coordinate-based} geometry, as expressed by the
type $A[0/t]$, in which the origin is fixed to be $(0,0)$. (It is sometimes said
that an affine space is a vector space that's forgotten its origin; we have captured
this in a type isomorphism.)

\exref{ex:type-iso} is a special case of this isomorphism; another instance is the
type of vector addition:
\begin{eqnarray*}
\forall t_1,t_2\mathord:\SynTransl{2}.\tyPrim{vec}{t_1}\to\tyPrim{vec}{t_2}\to\tyPrim{vec}{t_1+t_2}\\
\cong\quad
\forall t_2\mathord:\SynTransl{2}.\tyPrim{vec}{t_2}\to\tyPrim{vec}{t_2}\quad\cong\quad
\tyPrim{vec}{0}
\end{eqnarray*}

The maps associated with this isomorphism are as follows:
\newcommand{\trans}[3]{\mathrm{tr}_{#1}^{{#2}/{#3}}}
\begin{eqnarray*}
I = \lambda x. x\ [0]\ 0 & \text{ and } &
J = \lambda y. \Lambda t.\ \lambda v. \trans A v t(y)
\end{eqnarray*}
where $\mathrm{tr}$ is defined as follows:
\begin{eqnarray*}
\trans{1} v t(x) & = & x\\
\trans{A_1\times A_2} v t(p) & = & (\trans{A_1} v t(\pi_1\ p),\trans{A_2} v t(\pi_2\ p))\\
\trans{A_1 + A_2} v t(x) & = & \textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\trans{A_1} v t(y)); \textrm{inr}\ y.\textrm{inr}(\trans{A_2} v t(y))\\
\trans{A_1\to A_2} v t(f) & = & \lambda x. (\trans{A_2} v t(f\ (\trans{A_1}{-v} t(x))))\\
\trans{\tyPrim{vec}{e}} v t (x) & = & x + kv\text{ where }\vdash e \equiv k t + e' : \SynTransl{2}\text{ for some }k\in\setOfIntegers, t \text{ and }e'\\
\end{eqnarray*}
To show $I(J(y))$ is contextually equivalent to $y$ is an easy induction on $A$ that
makes use of the fact that the zero vector $0$ is an identity for vector addition. 
To show $J(I(x))$ is contextually equivalent to $x$ we must apply the abstraction theorem,
making use of the fact that $\mathrm{tr}$ is actually a term-based functional representation
of an instance of the relational interpretation:
\[
\rsem A {t \mapsto v} {} = \{ (x, \tmSem{\trans A v t (x)}) \mid x \in \tySem A\}.
\]

\paragraph{Scalings}
Now consider the group $\GL{1}$ of scalings. This time, we can treat the real-valued
argument to a function as a `unit of measure' with which to scale the result. The small
fly in the ointment is that the argument might be zero, and this cannot be used for scaling.
Nevertheless, we can obtain the following slightly more complicated isomorphism:
\begin{eqnarray*}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to A
& \cong &
A[1/s]\times \forall s\mathord:\SynGL{1}.A
\end{eqnarray*}
The type of $\mathrm{areaCircle}$ from the introduction is one instance:
\begin{eqnarray*}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to\tyPrim{real}{s\cdot s}\\
\quad\cong\quad \tyPrim{real}{1}\times(\forall s\mathord:\SynGL{1}.\tyPrim{real}{s\cdot s})\\
\quad\cong\quad \tyPrim{real}{1}\times 1\quad\cong\quad \tyPrim{real}{1}
\end{eqnarray*}
where the isomorphism $\forall s\mathord:\SynGL{1}.\tyPrim{real}{s\cdot s}\cong 1$ is
proved using the abstraction theorem.

%\begin{enumerate}
%\item Two cases: when we have a total ``division'' operation, and when
%  we don't.
%\item Type isomorphisms, based on Smith Normal Form
% \item Non-definability results, based on the non-compositional
%   relational environments
% \item Alteration of the non-definability results in the light of
%   addition of a square root operation
% \item Mention the variation we can obtain by changing some types,
%   e.g. the type of $\mathrm{abs}$. Also, the complex numbers example.
%\end{enumerate}
